Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome

### 2016-06-01 submission/Sze_Obesity_PeerJ_2016.Rmd
Code in the number of studies

@@

number_string <- c("one", "two", "three", "four", "five", "six", "seven", "eight", "nine")
```

@@

## Importance
As interest in the human microbiome grows there is an increasing number of studies that can be used to test numerous hypotheses across human populations. The hypothesis that variation in the gut microbiota can explain or be used to predict obesity status has received considerable attention and is frequently mentioned as an example for the role of the microbiome in human health. Here we assess this hypothesis using ten independent studies and find that although there is an association, it is smaller than can be detected by most microbiome studies. Furthermore, we directly tested the ability to predict obesity status based on the composition of an individual's microbiome and find that the median classification accuracy is between `r range(test_summary$accuracy)[1]` and `r range(test_summary$accuracy)[2]`%. This type of analysis can be used to design future studies and expanded to explore other hypotheses.

@@

***Literature Review and Study Inclusion***
We followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines to identify studies to include in our meta-analysis [@moher_preferred_2010]. A detailed description of our selection process and the exact search terms are provided in the Supplement and in figure 1. Briefly, we searched PubMed for original research studies that involved studying obesity and the human microbiome. The initial search yielded 187 studies. We identified ten additional studies that were not designed to explicitly test for an association between the microbiome and obesity. We then manually curated the 196 studies to select those studies that included BMI and sequence data. This yielded 10 eligible studies. An additional study was removed from our analysis because no individuals in the study had a BMI over 30. Among the final 10 studies, 3 were from identified from our PubMed search [@turnbaugh_core_2009; @zupancic_analysis_2012; @escobar_gut_2014], 5 were originally identified from the 9 studies that did not explicitly investigate obesity but included BMI data [@zeevi_personalized_2015; @ross_16s_2015; @goodrich_human_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012], and two datasets were used [@schubert_microbiome_2014; @baxter_microbiota-based_2016] because these publications did not specifically look for any metabolic or obesity conditions but had control populations and enabled us to help mitigate against publication biases associated with the bacterial microbiome and obesity. The ten studies are summarized in Tables 1 and 2.

@@

***Alpha diversity analysis***
We calculated the Shannon diversity index, richness, and Shannon evenness,  the relative abundance of *Bacteroidetes* and *Firmicutes*, and the ratio of their relative abundance (B:F) for each sample. Once we transformed each of the six alpha diversity metrics to make them normally distributed, we used a t-test to identify significant associations between the alpha diversity metric and whether an individual was obese for each of the ten studies. The B:F and the relative abundance of *Firmicutes* were not significantly associated with obesity in any study. We identified `r sum(alpha$p.value < 0.05)` P-values less than 0.05: `r number_string[sig_counts["sobs"]]` studies indicated obese individuals had a lower richness, `r number_string[sig_counts["shannon"]]` studies indicated a significantly lower diversity, `r number_string[sig_counts["shannoneven"]]` study indicated a significantly lower evenness, and `r number_string[sig_counts["bacteroidetes"]]` study indicated a significantly higher relative abundance of *Bacteroidetes* (Figures 2 and S1). These results largely match those of the Walters and Finucane re-analysis studies. Interestingly, although only `r number_string[sig_counts["shannon"]]` of the 10 studies observed the previously reported association between lower diversity and obesity, the other studies appeared to have the same trend, albeit the differences were not statistically significant. We used a random effects linear model to combine the studies using the study as the random effect and found statistical support for decreased richness, evenness, and diversity among obese individuals (all P<`r max(alpha_composite[alpha_composite$p <0.05, "p"])`). Although there was a significant relationship between these metrics and diversity and obesity status, the effect size was quite small. The obese individuals averaged `r alpha_effect_size["sobs"]`% lower richness, `r alpha_effect_size["shannoneven"]`% lower evenness, and `r alpha_effect_size["shannon"]`% lower diversity. There were no significant associations when we pooled the phylum-level metrics across studies. These results indicate that obese individuals do have a statistically significant lower diversity than non-obese individuals; however, it is questionable whether the difference is biologically significant.

@@

***Relative risk***
Building upon the alpha diversity analysis we calculated the relative risk of being obese based on whether an individual's alpha diversity metrics were below or above the median metric for that study. The results using relative risk largely matched those of using the untransformed alpha diversity data. Across the ten studies and six metrics, the only significant relative risk values were the richness, evenness, and diversity values from the Goodrich study (Figures 3 and S2). Again, although the relative risk values were not significant for other studies, the values tended to be above one. When we pooled the data using a random effects model, the relative risk associated with having a richness, evenness, or diversity below the median for the population was significantly associated with obesity (all P<`r max(relative_composite[relative_composite$p_value <0.05, "p_value"])`). The relative risks associated with alpha diversity were small. The relative risk of having a low richness was `r concatenate_rr(relative_composite["sobs",])`, low evenness was `r concatenate_rr(relative_composite["shannoneven",])`, and low diversity was `r concatenate_rr(relative_composite["shannon",])`. There were no significant difference in the phylum-level metrics. Again, the relative risk results indicate that individuals with a lower richness, evenness, or diversity are at statistically significant increased risk of being obese, it is questionable whether that risk is biologically or clinically relevant.

@@

***Development of a microbiome-based classifier of obesity***
The Walters re-analysis study suggested that it was possible to classify individuals as being obese or non-obese based on the composition of their microbiota. We repeated this analysis with additional datasets using OTU and genus-level phylotype data. For each study we developed a Random Forest machine learning model to classify individuals. Using ten-fold cross validation, the observed AUC values varied between `r min(otu$auc_cv)` and `r max(otu$auc_cv)` indicating a relatively poor ability to classify individuals (Figure 4A). So that we could test models on other datasets, we trained models using genus-level phylotype data for each dataset. The the observed AUC values for the models applied to the training datastes varied between `r min(genus$auc_cv)` and `r max(genus$auc_cv)`, again indicating a relatively poor ability to classify individuals from the original dataset (Figure 4B). For each model we identified the probability where the sum of the sensitivity and specificity was the highest. We then used this to identify a threshold to calculate the accuracy of the models when applied to the other nine datasets (Figure 5). Although there considerable variation in accuracy values for each model, the median accuracy for each model varied between `r min(test_summary$accuracy)` (Turnbaugh) and `r max(test_summary$accuracy)` (HMP) (median=`r median(test_summary$accuracy)`). When we considered the number of samples, balance of non-obese and obese individuals, and region within the 16S rRNA gene it was not possible to identify factors that predictably affected model performance. The ability to predict obesity status using the relative abundance of OTUs and genera in the communities is only marginally better than random. These results suggest that given the large diversity of microbiome compositions it is difficult to identify a taxonomic signal that can be associated with obesity.

>>>>

n <- length(unique(alpha$dataset))

@@

number_string <- c("one", "two", "three", "four", "five", "six", "seven", "eight", "nine", "ten")
```

@@

## Importance
As interest in the human microbiome grows there is an increasing number of studies that can be used to test numerous hypotheses across human populations. The hypothesis that variation in the gut microbiota can explain or be used to predict obesity status has received considerable attention and is frequently mentioned as an example for the role of the microbiome in human health. Here we assess this hypothesis using `r number_string[n]` independent studies and find that although there is an association, it is smaller than can be detected by most microbiome studies. Furthermore, we directly tested the ability to predict obesity status based on the composition of an individual's microbiome and find that the median classification accuracy is between `r range(test_summary$accuracy)[1]` and `r range(test_summary$accuracy)[2]`%. This type of analysis can be used to design future studies and expanded to explore other hypotheses.

@@

***Literature Review and Study Inclusion***
We followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines to identify studies to include in our meta-analysis [@moher_preferred_2010]. A detailed description of our selection process and the exact search terms are provided in the Supplement and in figure 1. Briefly, we searched PubMed for original research studies that involved studying obesity and the human microbiome. The initial search yielded 187 studies. We identified number_string[n] additional studies that were not designed to explicitly test for an association between the microbiome and obesity. We then manually curated the 196 studies to select those studies that included BMI and sequence data. This yielded 10 eligible studies. An additional study was removed from our analysis because no individuals in the study had a BMI over 30. Among the final 10 studies, 3 were from identified from our PubMed search [@turnbaugh_core_2009; @zupancic_analysis_2012; @escobar_gut_2014], 5 were originally identified from the 9 studies that did not explicitly investigate obesity but included BMI data [@zeevi_personalized_2015; @ross_16s_2015; @goodrich_human_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012], and two datasets were used [@schubert_microbiome_2014; @baxter_microbiota-based_2016] because these publications did not specifically look for any metabolic or obesity conditions but had control populations and enabled us to help mitigate against publication biases associated with the bacterial microbiome and obesity. The number_string[n] studies are summarized in Tables 1 and 2.

@@

***Alpha diversity analysis***
We calculated the Shannon diversity index, richness, and Shannon evenness,  the relative abundance of *Bacteroidetes* and *Firmicutes*, and the ratio of their relative abundance (B:F) for each sample. Once we transformed each of the six alpha diversity metrics to make them normally distributed, we used a t-test to identify significant associations between the alpha diversity metric and whether an individual was obese for each of the number_string[n] studies. The B:F and the relative abundance of *Firmicutes* were not significantly associated with obesity in any study. We identified `r sum(alpha$p.value < 0.05)` P-values less than 0.05: `r number_string[sig_counts["sobs"]]` studies indicated obese individuals had a lower richness, `r number_string[sig_counts["shannon"]]` studies indicated a significantly lower diversity, `r number_string[sig_counts["shannoneven"]]` study indicated a significantly lower evenness, and `r number_string[sig_counts["bacteroidetes"]]` study indicated a significantly higher relative abundance of *Bacteroidetes* (Figures 2 and S1). These results largely match those of the Walters and Finucane re-analysis studies. Interestingly, although only `r number_string[sig_counts["shannon"]]` of the 10 studies observed the previously reported association between lower diversity and obesity, the other studies appeared to have the same trend, albeit the differences were not statistically significant. We used a random effects linear model to combine the studies using the study as the random effect and found statistical support for decreased richness, evenness, and diversity among obese individuals (all P<`r max(alpha_composite[alpha_composite$p <0.05, "p"])`). Although there was a significant relationship between these metrics and diversity and obesity status, the effect size was quite small. The obese individuals averaged `r alpha_effect_size["sobs"]`% lower richness, `r alpha_effect_size["shannoneven"]`% lower evenness, and `r alpha_effect_size["shannon"]`% lower diversity. There were no significant associations when we pooled the phylum-level metrics across studies. These results indicate that obese individuals do have a statistically significant lower diversity than non-obese individuals; however, it is questionable whether the difference is biologically significant.

@@

***Relative risk***
Building upon the alpha diversity analysis we calculated the relative risk of being obese based on whether an individual's alpha diversity metrics were below or above the median metric for that study. The results using relative risk largely matched those of using the untransformed alpha diversity data. Across the number_string[n] studies and six metrics, the only significant relative risk values were the richness, evenness, and diversity values from the Goodrich study (Figures 3 and S2). Again, although the relative risk values were not significant for other studies, the values tended to be above one. When we pooled the data using a random effects model, the relative risk associated with having a richness, evenness, or diversity below the median for the population was significantly associated with obesity (all P<`r max(relative_composite[relative_composite$p_value <0.05, "p_value"])`). The relative risks associated with alpha diversity were small. The relative risk of having a low richness was `r concatenate_rr(relative_composite["sobs",])`, low evenness was `r concatenate_rr(relative_composite["shannoneven",])`, and low diversity was `r concatenate_rr(relative_composite["shannon",])`. There were no significant difference in the phylum-level metrics. Again, the relative risk results indicate that individuals with a lower richness, evenness, or diversity are at statistically significant increased risk of being obese, it is questionable whether that risk is biologically or clinically relevant.

@@

***Development of a microbiome-based classifier of obesity***
The Walters re-analysis study suggested that it was possible to classify individuals as being obese or non-obese based on the composition of their microbiota. We repeated this analysis with additional datasets using OTU and genus-level phylotype data. For each study we developed a Random Forest machine learning model to classify individuals. Using ten-fold cross validation, the observed AUC values varied between `r min(otu$auc_cv)` and `r max(otu$auc_cv)` indicating a relatively poor ability to classify individuals (Figure 4A). So that we could test models on other datasets, we trained models using genus-level phylotype data for each dataset. The the observed AUC values for the models applied to the training datastes varied between `r min(genus$auc_cv)` and `r max(genus$auc_cv)`, again indicating a relatively poor ability to classify individuals from the original dataset (Figure 4B). For each model we identified the probability where the sum of the sensitivity and specificity was the highest. We then used this to identify a threshold to calculate the accuracy of the models when applied to the other number_string[n-1] datasets (Figure 5). Although there considerable variation in accuracy values for each model, the median accuracy for each model varied between `r min(test_summary$accuracy)` (Turnbaugh) and `r max(test_summary$accuracy)` (HMP) (median=`r median(test_summary$accuracy)`). When we considered the number of samples, balance of non-obese and obese individuals, and region within the 16S rRNA gene it was not possible to identify factors that predictably affected model performance. The ability to predict obesity status using the relative abundance of OTUs and genera in the communities is only marginally better than random. These results suggest that given the large diversity of microbiome compositions it is difficult to identify a taxonomic signal that can be associated with obesity.
